---
title: "Getting Started with DynamicAlphaHull"
author: "Yuxuan Wang"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with DynamicAlphaHull}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Introduction

**DynamicAlphaHull** is an R package for building species geographic
ranges using alpha-hull methods with dynamic alpha parameter selection.
It provides:

- Efficient parallel processing of multiple species
- Automatic alpha parameter optimization based on point density
- Integration with 1-degree grid systems for biogeographic analysis
- Vectorized geometry operations for performance
- Reproducible workflows with deterministic random number generation

This vignette demonstrates the basic workflow using the included
`rosales_example` dataset.

## Installation

```{r install, eval = FALSE}
# Install from GitHub
# devtools::install_github("yourname/DynamicAlphaHull")

# Load the package
library(DynamicAlphaHull)
```

```{r load, message = FALSE}
library(DynamicAlphaHull)
library(data.table)
library(terra)
```

## Example Data

The package includes `rosales_example`, a dataset of 31,742 occurrence
records for 100 Rosales species:

```{r data}
data(rosales_example)

# Dataset structure
str(rosales_example)

# Summary statistics
cat("Total records:", nrow(rosales_example), "\n")
cat("Number of species:", uniqueN(rosales_example$accepted_name), "\n")
cat("Data sources:", paste(unique(rosales_example$data_source), collapse = ", "), "\n")
```

### Species Distribution

Let's look at the species with the most occurrence records:

```{r species_summary}
species_counts <- rosales_example[, .N, by = accepted_name][order(-N)]
head(species_counts, 10)
```

### Geographic Coverage

```{r geographic_coverage}
# Coordinate ranges
cat("Longitude range:", range(rosales_example$longitude), "\n")
cat("Latitude range:", range(rosales_example$latitude), "\n")

# Records per data source
rosales_example[, .N, by = data_source]
```

## Basic Workflow

### Step 1: Prepare Input Data

Your occurrence data should be a `data.table` with columns: -
`accepted_name` (or mapped via column mapping): species identifier -
`longitude`: decimal degrees - `latitude`: decimal degrees -
`data_source`: optional, source identifier

```{r inspect_data}
# Look at a few records for one species
rosales_example[accepted_name == "Urtica urens"][1:5]
```

### Step 2: Load Reference Grid

The package includes a 1-degree global grid for range discretization:

```{r load_grid}
# Load the included 1-degree grid
data(grid_1deg)
grid <- terra::unwrap(grid_1deg)

# Grid properties
cat("Grid cells:", terra::nrow(grid), "\n")
cat("CRS:", terra::crs(grid, describe = TRUE)$name, "\n")
cat("Coverage:", terra::ext(grid)[1:4], "\n")

# Create range context
context <- new_range_context(grid)
```

The `grid_1deg` dataset contains 19,196 cells covering the globe from
56°S to 84°N. Each cell has a unique `grid_id` for efficient range
representation.

### Step 3: Build Alpha-Hull Ranges

Use `build_family_ranges()` to compute ranges for all species in
parallel:

```{r build_ranges, eval = FALSE}
# Set seed for reproducibility
set.seed(42)

# Build ranges with 4 workers
result <- build_family_ranges(
  rosales_example,
  context,
  workers = 4L,
  alpha_timeout_seconds = 300,
  progress = TRUE
)
```

**Parameters:** - `workers`: Number of parallel workers (default: 4) -
`alpha_timeout_seconds`: Timeout per species in seconds (default: 300) -
`progress`: Show progress messages (default: FALSE) - `progress_file`:
Optional file path to log progress

### Step 4: Examine Results

The result contains two components:

```{r examine_results, eval = FALSE}
# Summary table: one row per species
names(result$summary)

# Success rate
cat("Successful species:", sum(result$summary$success), "/",
    nrow(result$summary), "\n")

# Range areas
summary(result$summary$range_area_km2)

# Processing time
summary(result$summary$hull_elapsed_seconds)
```

**Summary columns:** - `species`: Species name - `point_count`: Number
of occurrence records - `method`: Method used ("direct" for \<50 points,
"alpha_hull" otherwise) - `success`: Whether range building succeeded -
`grid_count`: Number of 1-degree grid cells in the range -
`range_area_km2`: Total range area in square kilometers -
`hull_elapsed_seconds`: Processing time for this species -
`error_message`: Error details if `success == FALSE`

```{r grid_distribution, eval = FALSE}
# Grid distribution: grid cell IDs for each species
str(result$grid_distribution)

# Example: cells for one species
head(result$grid_distribution[species == "Urtica urens"])
```

## Performance Considerations

### Species Filtering

For large datasets, consider filtering species by occurrence count:

```{r filtering, eval = FALSE}
# Keep species with 50-3000 occurrences
species_counts <- rosales_example[, .N, by = accepted_name]
moderate_species <- species_counts[N >= 50 & N < 3000, accepted_name]

filtered_data <- rosales_example[accepted_name %in% moderate_species]

# This avoids timeout issues with very large species
```

### Parallel Configuration

Adjust workers based on your system:

```{r parallel_config, eval = FALSE}
# Conservative: use half of available cores
workers <- max(1L, parallel::detectCores() %/% 2L)

# Aggressive: use most cores (leave 1-2 for system)
workers <- max(1L, parallel::detectCores() - 1L)
```

### Timeout Settings

Species with many points (\>3000) may need longer timeouts:

```{r timeout_config, eval = FALSE}
# Default timeout
result <- build_family_ranges(pts, context, alpha_timeout_seconds = 300)

# Longer timeout for large species
result <- build_family_ranges(pts, context, alpha_timeout_seconds = 600)
```

## Reproducibility

For reproducible results, always set a seed before calling
`build_family_ranges()`:

```{r reproducibility, eval = FALSE}
set.seed(42)
result1 <- build_family_ranges(rosales_example, context, workers = 4L)

set.seed(42)
result2 <- build_family_ranges(rosales_example, context, workers = 4L)

# Results should be identical
identical(result1$summary, result2$summary)
identical(result1$grid_distribution, result2$grid_distribution)
```

This works because: 1. `furrr_options(seed = TRUE)` derives per-task
seeds from the master seed 2. Results are independent of worker count
given the same master seed 3. The jitter mechanism in alpha-hull
triangulation is deterministic

## Example Analysis

Here's a complete example analyzing the included dataset:

```{r complete_example, eval = FALSE}
library(DynamicAlphaHull)
library(data.table)
library(ggplot2)

# Load data
data(rosales_example)

# Load grid and create context
data(grid_1deg)
grid <- terra::unwrap(grid_1deg)
context <- new_range_context(grid)

# Build ranges
set.seed(42)
t0 <- Sys.time()
result <- build_family_ranges(
  rosales_example,
  context,
  workers = 8L,
  alpha_timeout_seconds = 300,
  progress = TRUE
)
elapsed <- difftime(Sys.time(), t0, units = "secs")

# Summary statistics
cat("Processing time:", round(elapsed, 1), "seconds\n")
cat("Success rate:", sum(result$summary$success), "/",
    nrow(result$summary), "\n")
cat("Average per species:", round(elapsed / nrow(result$summary), 2), "s\n")

# Analyze relationship between points and range size
setDT(result$summary)
ggplot(result$summary[success == TRUE],
       aes(x = point_count, y = grid_count)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "loess") +
  scale_x_log10() +
  scale_y_log10() +
  labs(
    x = "Number of occurrence records",
    y = "Number of 1-degree grid cells",
    title = "Relationship between occurrences and range size"
  ) +
  theme_minimal()

# Top 10 largest ranges
result$summary[order(-range_area_km2)][1:10,
  .(accepted_name = species, point_count, grid_count, range_area_km2)]
```

## Next Steps

- See `?build_family_ranges` for detailed parameter documentation
- Explore the validation scripts in `validation/` for production
  workflows
- Check out performance benchmarks in `validation/output/`

## Session Info

```{r session_info}
sessionInfo()
```
