---
title: "Getting Started with DynamicAlphaHull"
author: "Yuxuan Wang"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with DynamicAlphaHull}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  eval = TRUE
)
```

## Introduction

**DynamicAlphaHull** constructs species geographic range polygons from occurrence coordinates using an adaptive alpha-hull algorithm. The package automatically selects an appropriate alpha parameter to create tight-fitting range boundaries while maintaining connectivity.

## Basic Usage

```{r basic}
library(DynamicAlphaHull)

# Create example occurrence points
points <- data.frame(
  Longitude = c(116.0, 116.5, 116.4, 116.1, 115.8, 116.6),
  Latitude = c(39.0, 39.0, 39.4, 39.3, 39.1, 39.2)
)

# Build a range polygon
result <- getDynamicAlphaHull(
  points,
  buff = 10000,
  clipToCoast = "no"
)

# View the result
plot(result$hull, col = "lightblue", border = "blue", main = "Species Range")
points(points$Longitude, points$Latitude, pch = 16, cex = 1.5, col = "red")
```

The function returns:

- `hull`: A `terra::SpatVector` polygon
- `alpha`: The selected alpha value

```{r inspect}
cat("Selected alpha:", result$alpha, "\n")
```

## Key Parameters

**buff**: Buffer distance in meters (default: 10,000m)

**fraction**: Minimum proportion of points to include (default: 0.95)

**partCount**: Maximum disconnected parts (default: 3)

**clipToCoast**: Clipping option - `"no"`, `"terrestrial"`, or `"aquatic"`

## Working with Real Data

The package includes example occurrence data:

```{r example_data}
data(rosales_example)
head(rosales_example, 3)
```

Build a range for one species:

```{r species_range}
# Select one species
species_data <- rosales_example[rosales_example$accepted_name == "Urtica urens", ]

cat("Occurrences:", nrow(species_data), "\n")

# Build range (using data.frame to avoid data.table issues)
species_df <- as.data.frame(species_data)

range_result <- getDynamicAlphaHull(
  species_df,
  coordHeaders = c("longitude", "latitude"),
  buff = 10000,
  clipToCoast = "terrestrial"
)

cat("Alpha selected:", range_result$alpha, "\n")
```

Visualize the range:

```{r visualize}
plot(range_result$hull, 
     col = adjustcolor("forestgreen", alpha = 0.3),
     border = "darkgreen", lwd = 2,
     main = "Urtica urens Range")
points(species_df$longitude, species_df$latitude,
       pch = 16, cex = 0.4, col = "darkred")
```

## Calculating Range Area

```{r area}
library(terra)

# Project to equal-area for accurate measurement
range_proj <- project(range_result$hull, "EPSG:6933")
area_km2 <- expanse(range_proj, unit = "km")

cat("Range area:", formatC(area_km2, format = "f", big.mark = ",", digits = 0), "km²\n")
```

## Grid Representation

Convert ranges to grid cells for comparative analysis:

```{r grid}
data(grid_1deg)
grid <- unwrap(grid_1deg)

# Find intersecting cells
pairs <- relate(grid, range_result$hull, relation = "intersects", pairs = TRUE)

if (is.matrix(pairs) && nrow(pairs) > 0) {
  occupied_cells <- sort(unique(grid$grid_id[pairs[, 1]]))
  cat("Occupied grid cells:", length(occupied_cells), "\n")
  cat("First 10 cells:", paste(head(occupied_cells, 10), collapse = ", "), "\n")
}
```

## Processing Multiple Species

```{r batch, eval = FALSE}
# Get species names
species_list <- unique(rosales_example$accepted_name)

# Process each species
results <- list()
for (sp in species_list[1:5]) {  # First 5 species as example
  sp_data <- rosales_example[rosales_example$accepted_name == sp, ]
  
  if (nrow(sp_data) >= 10) {  # Minimum points
    result <- tryCatch({
      getDynamicAlphaHull(
        as.data.frame(sp_data),
        coordHeaders = c("longitude", "latitude"),
        clipToCoast = "terrestrial"
      )
    }, error = function(e) NULL)
    
    if (!is.null(result)) {
      results[[sp]] <- result
    }
  }
}

cat("Successfully processed:", length(results), "species\n")
```

## Reproducibility

Set a seed for reproducible results:

```{r seed}
set.seed(1234)
result1 <- getDynamicAlphaHull(points, clipToCoast = "no")

set.seed(1234)
result2 <- getDynamicAlphaHull(points, clipToCoast = "no")

identical(result1$alpha, result2$alpha)
```

## Further Information

- `?getDynamicAlphaHull` - Full parameter documentation
- `?rosales_example` - Example dataset details  
- `?grid_1deg` - Reference grid documentation

## Session Information

```{r session}
sessionInfo()
```
