---
title: "Getting Started with DynamicAlphaHull"
author: "Yuxuan Wang"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with DynamicAlphaHull}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Introduction

**DynamicAlphaHull** constructs species geographic range polygons from
occurrence coordinates using an adaptive alpha-hull algorithm. The
package automatically selects an appropriate alpha parameter to balance
two goals: tightly fitting the occurrence points while maintaining a
connected range polygon.

This vignette demonstrates the core workflow using the main function
`getDynamicAlphaHull()`.

## Installation

```{r install, eval = FALSE}
# Install from GitHub
# remotes::install_github("YourUsername/DynamicAlphaHull")

# Load the package
library(DynamicAlphaHull)
```

```{r load, echo = FALSE}
library(DynamicAlphaHull)
```

## Quick Start

The simplest workflow requires only occurrence coordinates:

```{r quickstart}
# Create a small occurrence dataset
points <- data.frame(
  Longitude = c(116.0, 116.5, 116.4, 116.1, 115.8, 116.6),
  Latitude = c(39.0, 39.0, 39.4, 39.3, 39.1, 39.2)
)

# Build a range polygon
result <- getDynamicAlphaHull(
  points,
  buff = 10000,        # 10km buffer
  clipToCoast = "no"   # don't clip to coastline
)

# Plot the result
plot(result$hull, col = "lightblue", border = "blue")
points(points$Longitude, points$Latitude, pch = 16, cex = 1.5)
```

The function returns a list with two elements:

- `hull`: A `terra::SpatVector` polygon representing the species range
- `alpha`: The selected alpha value (as a character label)

```{r quickstart_inspect}
cat("Selected alpha:", result$alpha, "\n")
cat("Range is a", class(result$hull)[1], "\n")
```

## Understanding Parameters

### Core Parameters

**alpha**: Controls how tightly the polygon wraps around points. The
algorithm starts at `initialAlpha` and increases by `alphaIncrement`
until it finds a suitable value.

**fraction**: Minimum proportion of occurrences that must fall within
the polygon (default: 0.95). Higher values produce more inclusive
ranges.

**partCount**: Maximum number of disconnected polygon parts allowed
(default: 3). This prevents highly fragmented ranges.

**buff**: Buffer distance in meters applied to the final polygon
(default: 10,000m). This accounts for spatial uncertainty in occurrence
records.

### Coastline Clipping

**clipToCoast**: Controls whether to clip the range to land or water:

- `"no"`: No clipping (default for quick examples)
- `"terrestrial"`: Clip to land (for terrestrial species)
- `"aquatic"`: Clip to water bodies (for aquatic species)

```{r coastline_demo, eval = FALSE}
# Terrestrial species - clip to land
terrestrial_range <- getDynamicAlphaHull(
  points,
  clipToCoast = "terrestrial"
)
```

The package includes an offline Natural Earth 1:50m land polygon, so no
internet connection is required.

## Working with Example Data

The package includes occurrence data for 100 Rosales species:

```{r example_data}
data(rosales_example)
head(rosales_example)
```

Each record contains:

- `accepted_name`: Species name
- `longitude`, `latitude`: Coordinates in decimal degrees
- `data_source`: Data origin (WCVP or GBIF)

### Single Species Example

Let's build a range for one species:

```{r single_species}
library(data.table)
setDT(rosales_example)

# Select a species with moderate occurrence count
species_name <- "Rosa canina"
species_data <- rosales_example[accepted_name == species_name]

cat("Building range for", species_name, "\n")
cat("Occurrence records:", nrow(species_data), "\n")

# Build the range
range_result <- getDynamicAlphaHull(
  species_data,
  fraction = 0.95,
  partCount = 3,
  buff = 10000,
  initialAlpha = 3,
  coordHeaders = c("longitude", "latitude"),
  clipToCoast = "terrestrial"
)

cat("Selected alpha:", range_result$alpha, "\n")
```

### Visualizing the Range

```{r visualize}
# Plot the range polygon
plot(range_result$hull, 
     col = adjustcolor("forestgreen", alpha = 0.3),
     border = "darkgreen",
     lwd = 2,
     main = paste("Geographic Range:", species_name))

# Add occurrence points
points(species_data$longitude, 
       species_data$latitude,
       pch = 16, 
       cex = 0.3, 
       col = "darkred")
```

### Calculating Range Size

```{r range_size}
library(terra)

# Project to equal-area projection for accurate area calculation
range_proj <- project(range_result$hull, "EPSG:6933")
area_km2 <- expanse(range_proj, unit = "km")

cat("Range area:", round(area_km2), "km²\n")
```

## Grid-Based Range Representation

For comparative analyses across many species, it's often useful to
represent ranges as sets of grid cells rather than complex polygons. The
package includes a 1-degree global reference grid:

```{r grid_setup}
data(grid_1deg)
grid <- unwrap(grid_1deg)

cat("Grid cells:", nrow(grid), "\n")
head(grid)
```

### Finding Occupied Grid Cells

```{r grid_intersection}
# Find which grid cells the range intersects
pairs <- relate(grid, range_result$hull, relation = "intersects", pairs = TRUE)

if (is.matrix(pairs) && nrow(pairs) > 0) {
  occupied_cells <- unique(grid$grid_id[pairs[, 1]])
  cat("Grid cells occupied:", length(occupied_cells), "\n")
  cat("First 10 cell IDs:", head(occupied_cells, 10), "\n")
}
```

This compact representation is useful for:

- Efficiently storing ranges for many species
- Computing range overlaps
- Calculating diversity metrics
- Comparing range sizes across species

### Visualizing Grid Representation

```{r grid_viz, eval = FALSE}
# Extract occupied grid cells
occupied_grid <- grid[grid$grid_id %in% occupied_cells, ]

# Plot both representations
plot(range_result$hull, 
     col = adjustcolor("blue", alpha = 0.3),
     border = "blue",
     main = "Range Polygon vs Grid Representation")
plot(occupied_grid, 
     col = adjustcolor("red", alpha = 0.2),
     border = "red",
     add = TRUE)
```

## Processing Multiple Species

For batch processing, you can loop over species or use parallel
processing:

```{r batch_basic, eval = FALSE}
# Get species with moderate occurrence counts
counts <- rosales_example[, .N, by = accepted_name]
target_species <- counts[N >= 50 & N <= 500, accepted_name]

# Process each species
results_list <- list()
for (sp in target_species) {
  sp_data <- rosales_example[accepted_name == sp]
  
  result <- tryCatch({
    getDynamicAlphaHull(
      sp_data,
      coordHeaders = c("longitude", "latitude"),
      clipToCoast = "terrestrial"
    )
  }, error = function(e) NULL)
  
  if (!is.null(result)) {
    results_list[[sp]] <- result
  }
}
```

### Parallel Processing

For larger datasets, use parallel processing to speed up computation:

```{r parallel, eval = FALSE}
library(future)
library(furrr)

# Set up parallel processing
plan(multisession, workers = 4)

# Process species in parallel
results <- target_species |>
  furrr::future_map(
    function(sp) {
      sp_data <- rosales_example[accepted_name == sp]
      getDynamicAlphaHull(
        sp_data,
        coordHeaders = c("longitude", "latitude"),
        clipToCoast = "terrestrial"
      )
    },
    .options = furrr_options(seed = TRUE)
  )

names(results) <- target_species
```

## Special Cases

### Small Occurrence Datasets

Species with very few occurrences (\<10 points) may not produce
meaningful alpha-hulls. Consider using a minimum convex polygon or
buffered points instead:

```{r small_dataset, eval = FALSE}
small_species <- rosales_example[accepted_name == "Rosa minutifolia"]
cat("Occurrences:", nrow(small_species), "\n")

# For very small datasets, the algorithm will fall back to 
# a buffered minimum convex hull automatically
```

### Antimeridian Handling

Species with ranges crossing the 180° longitude line are handled
correctly through `terra`'s geometry validation:

```{r antimeridian, eval = FALSE}
# Example: Pacific-spanning species
# The package automatically handles coordinate wrapping
```

### Large Occurrence Datasets

For species with \>5000 occurrences, consider:

- Spatial thinning to reduce computation time
- Increasing `alphaCap` if the algorithm fails to converge
- Using a coarser initial alpha value

```{r large_dataset, eval = FALSE}
large_species <- rosales_example[accepted_name == "Rosa canina"]

# Spatial thinning example (if needed)
# thin_data <- large_species[sample(.N, min(.N, 3000))]

# Or adjust algorithm parameters
result <- getDynamicAlphaHull(
  large_species,
  initialAlpha = 5,      # Start with larger alpha
  alphaIncrement = 2,    # Increase faster
  alphaCap = 600,        # Allow higher maximum
  coordHeaders = c("longitude", "latitude")
)
```

## Performance Tips

1.  **Occurrence filtering**: Remove duplicate coordinates and spatial
    outliers before processing
2.  **Parallel processing**: Use `future` and `furrr` for batch
    processing
3.  **Parameter tuning**: Adjust `initialAlpha` and `alphaIncrement`
    based on your data
4.  **Coastline clipping**: Use `clipToCoast = "no"` during development,
    add clipping for final results

## Reproducibility

Set a random seed for reproducible results, as the algorithm uses
randomization when handling collinear points:

```{r reproducibility}
# First run
set.seed(7892)
result1 <- getDynamicAlphaHull(points, clipToCoast = "no")

# Second run with same seed
set.seed(7892)
result2 <- getDynamicAlphaHull(points, clipToCoast = "no")

# Results are identical
identical(result1$alpha, result2$alpha)
```

## Further Reading

- `?getDynamicAlphaHull` - Complete parameter documentation
- `?rosales_example` - Example occurrence dataset details
- `?grid_1deg` - Reference grid documentation
- `?loadWorldMap` - Coastline data access

## Session Info

```{r session_info}
sessionInfo()
```
